---
title: "CAPM & Beta Estimation (Chapter 8 in Bali)"
format: html
toc: true
editor: visual
editor_options: 
  chunk_output_type: console
execute: 
  eval: false
---

In most cases, market beta for any given period t is estimated using **only data from periods prior to and including period t**. A common approach is to estimate regression (8.3) using **one year’s worth of daily excess return** data.

minimum obs requirement: 200 data points

Correlation:

There is a substantial common component between all of the measures of beta. The correlations between measures that use a longer measurement period are higher than those based on short measurement periods, potentially indicating that a **longer measurement period provides stable and therefore more accurate measurement**.

Persistence:

The use of longer measurement periods results in more accurate measures of beta. In addition, beta seems to be most accurately mea- sured using daily return data.

```{r}
library(tidyverse)
library(duckdb)
library(scales)
library(slider)
library(furrr)
source(here::here('R/my_functions.R'))
```

```{r}
tidy_finance <- dbConnect(duckdb(), here::here('data/tidy_finance'))
crsp_monthly <- tbl(tidy_finance, "crsp_monthly") |>
  select(permno, month, industry, ret_excess) |>
  collect()
factors_ff3_monthly <- tbl(tidy_finance, "factors_ff3_monthly") |>
  select(month, mkt_excess) |>
  collect()
crsp_monthly <- crsp_monthly |>
  left_join(factors_ff3_monthly, join_by(month))
```

```{r}
estimate_capm <- function(data, min_obs = 1) {
  if (nrow(data) < min_obs) {
    beta <- NA_real_
  } else {
    fit <- lm(ret_excess ~ mkt_excess, data = data)
    beta <- as.numeric(coefficients(fit)[2])
  }
  return(beta)
}

roll_capm_estimation <- function(data, months, min_obs = 1){
  data <- data %>% arrange(month)
  betas <- slide_period_vec(
    .x = data, 
    .i = data$month, 
    .period = 'month',
    .before = months - 1, 
    .complete = F, 
    .f = ~estimate_capm(., min_obs)
  )
  
  tibble(
    month = data$month, 
    beta = betas
  )
}

a <- crsp_monthly %>% 
  filter(permno == 14593) %>% 
  roll_capm_estimation(months = 60, min_obs = 48)
```

## Parallelization

```{r}
n_cores <- availableCores() - 1
plan(multisession, workers = n_cores)
beta_monthly <- crsp_monthly %>% 
  nest(data = c(month, ret_excess, mkt_excess)) %>% 
  mutate(beta = future_map(
    data, ~ roll_capm_estimation(., months = 60, min_obs = 48)
  )) |>
  unnest(beta) |>
  select(permno, month, beta_monthly = beta) |>
  drop_na()

dbWriteTable(tidy_finance, 'beta', beta_monthly, overwrite = T)
```

## ts-average beta across industries

```{r}
beta_monthly %>% 
  left_join(crsp_monthly, join_by(permno, month)) %>% 
  group_by(industry, permno) %>% 
  summarise(beta = mean(beta_monthly), .groups = 'drop') %>% 
  ggplot(aes(x = reorder(industry, beta, FUN = median), y = beta)) + 
  geom_boxplot() + 
  coord_flip() + 
  theme_hierarchy()
```

## cs beta over time

```{r}
beta_monthly %>% 
  group_by(month) %>% 
  reframe(quantile = seq(0.1, 0.9, 0.1), 
          x = quantile(beta_monthly, quantile)) %>% 
  ggplot(aes(month, x, color = as.factor(quantile*100))) +  
  geom_line() + 
  theme_hierarchy() + 
  ggsci::scale_color_bmj()
```
