---
title: "Fama-MacBeth Regressions"
format: html
toc: true
editor: visual
editor_options: 
  chunk_output_type: console
execute: 
  eval: false
---

Use: to estimate the risk premium associated with risk factors

The basic idea of the regression approach is to project asset returns on factor exposures or characteristics that resemble exposure to a risk factor in the cross-section **in each time period**. Then, in the second step, the estimates are aggregated across time to test if a risk factor is priced.

$$
\begin{aligned}r_{i,t+1} = \alpha_i + \lambda^{M}_t \beta^M_{i,t}  + \lambda^{SMB}_t \beta^{SMB}_{i,t} + \lambda^{HML}_t \beta^{HML}_{i,t} + \epsilon_{i,t}.\end{aligned}
$$

#### 2-Step Procedure

Step 0: estimate $\beta$ in TS regression or simply use characteristics

Step 1: estimate $\lambda$ cross-sectionally period by period

Step 2: time-series mean + sd for HT

## Cross-sectional regression

estimate the risk premium from 3 characteristics: size (log market cap), bm ratio, and market beta

```{r}
pacman::p_load(tidyverse, duckdb, sandwich, broom)

tidy_finance <- dbConnect(duckdb(), 'data/tidy_finance')
crsp_monthly <- tbl(tidy_finance, "crsp_monthly") |>
  select(permno, gvkey, month, ret_excess, mktcap) |>
  collect()

compustat <- tbl(tidy_finance, "compustat") |>
  select(datadate, gvkey, be) |>
  collect()

beta <- tbl(tidy_finance, "beta") |>
  select(month, permno, beta_monthly) |>
  collect()
```

```{r}
characteristics <- compustat |>
  mutate(month = floor_date(ymd(datadate), "month")) |>
  left_join(crsp_monthly, join_by(gvkey, month)) |>
  left_join(beta, join_by(permno, month)) |>
  transmute(gvkey,
    bm = be / mktcap,
    log_mktcap = log(mktcap),
    beta = beta_monthly,
    sorting_date = month %m+% months(6)
  )

data_fama_macbeth <- crsp_monthly |>
  left_join(characteristics, join_by(gvkey, month == sorting_date)) |>
  group_by(permno) |>
  arrange(month) |>
  fill(c(beta, bm, log_mktcap), .direction = "down") |>
  ungroup() |>
  left_join(crsp_monthly |>
    select(permno, month, ret_excess_lead = ret_excess) |>
    mutate(month = month %m-% months(1)),
  join_by(permno, month)
  ) |>
  select(permno, month, ret_excess_lead, beta, log_mktcap, bm) |>
  drop_na()
```

Why are the chars lagged by 7 months instead of 1 month?

At each $t$, run the cross-sectional regression:

```{r}
risk_premiums <- data_fama_macbeth |>
  nest(data = -month) |>
  mutate(estimates = map(
    data,
    ~ tidy(lm(ret_excess_lead ~ beta + log_mktcap + bm, data = .x))
  )) |>
  unnest(estimates)
```

## Time-series Aggregation

note that the denominator is the sd of the mean.

```{r}
price_of_risk <- risk_premiums |>
  group_by(factor = term) |>
  summarize(
    risk_premium = mean(estimate),
    t_statistic = mean(estimate) / sd(estimate) * sqrt(n())
  )
```

Newey-West SD via regression (to constant)

```{r}
regressions_for_newey_west <- risk_premiums %>% 
  select(month, factor = term, estimate) %>% 
  nest(data = -factor) %>% 
  mutate(model = map(data, ~lm(estimate ~ 1, .)), 
         mean = map(model, tidy))

price_of_risk_newey_west <- regressions_for_newey_west %>% 
  mutate(newey_west_se = map_dbl(model, ~sqrt(NeweyWest(.)))) %>% 
  unnest(mean) %>% 
  mutate(t_statistic_newey_west = estimate / newey_west_se) %>% 
  select(factor, risk_premium = estimate,
    t_statistic_newey_west)

left_join(price_of_risk,
  price_of_risk_newey_west |>
    select(factor, t_statistic_newey_west),
  join_by(factor)
)

```

```{r}
tidyfinance::estimate_fama_macbeth(
  data = data_fama_macbeth, 
  model = "ret_excess_lead ~ beta + bm + log_mktcap",
  vcov = "newey-west", 
  data_options = tidyfinance::data_options(date = 'month')
)
```
